DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2024-03-29 submission/manuscript.tex
3rd review edits

Word count: 7484

@@

At the time of writing of this manuscript
(Nov 21, 2023),
DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies, encompassing close to 100 k (99 474) different species. It is curated from OpenTree's phylogenetic database, the Phylesystem, an open database of expert and peer-reviewed phylogenetic knowledge with rich metadata and a wide taxonomic scope (McTavish et al., 2015). We expect DateLife's database to largely overlap with OpenTree's phylogenetic database taxonomic coverage, where Chordata and Embryophyta are nearly fully sampled. In contrast, Bacteria, Fungi, Nematoda, and Insecta, currently present a large gap between the number of named species and what has phylogenetic information in OpenTree's synthetic tree. It is likely that users working with the former groups will get results from a DateLife analysis. If none of the user's species are found, the software will indicate the lack of age data for the queried taxa in the database.

@@

Figure 2. DateLife results of an analysis of a small sample of 6 bird species within the Passeriformes (a, b). Processed species names were found across 9 chronograms within 6 independent studies (c; Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}). This revealed 28 source age data points for the queried species names (e; Table 1). Summarized age data (f; Table 2) was used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree v13.4 (d), resulting in the chronogram of summary source ages shown in (g).

Figure 3. Median summary chronogram resulting from a Datelife analysis of bird species within the family Fringillidae. For visualization purposes, we are showing a portion of the final median summary chronogram encompassing 57 species out of the 289 total included in the analysis. The complete final chronogram is available as Supplementary Figure S3. The starting tree topology (Supplementary Fig. S2) has 289 tip species and 253 nodes; DateLife revealed age data for 194 of these nodes from at least one published chronogram. In total, 19 different chronograms from 13 different studies contributed 818 age data points, which were summarized to obtain a single value for each one of the 194 nodes with age data. From the 194 summary ages available, 21 were discarded and not used as calibrations (asterisk, *), because they were older than a parent node or younger than a descendant node; the remaining 173 summary ages were used as secondary calibrations (forward slash, /) with the Branch Length Adjuster (BLADJ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).

>>>>

Word count: 7530

@@

At the time of writing of this manuscript
(Mar 08, 2024),
DateLife's chronogram database latest version consist of 253 chronograms published in 187 studies, encompassing 99 474 species. It is curated from OpenTree's phylogenetic database, the Phylesystem, an open database of expert and peer-reviewed phylogenetic knowledge with rich metadata and a wide taxonomic scope (McTavish et al., 2015). We expect DateLife's database to largely overlap with OpenTree's phylogenetic database taxonomic coverage, where Chordata and Embryophyta are nearly fully sampled. In contrast, Bacteria, Fungi, Nematoda, and Insecta, currently present a large gap between the number of named species and what has phylogenetic information in OpenTree's synthetic tree. It is likely that users working with the former groups will get results from a DateLife analysis. If none of the user's species are found, the software will indicate the lack of age data for the queried taxa in the database.

@@

Figure 2. DateLife results of an analysis of a small sample of 6 bird species within the Passeriformes (a, b). Processed species names were found across 9 chronograms within 6 independent studies (c; Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}). This revealed 28 source age data points for the queried species names (e; Table 1). Summarized age data (f; Table 2) was used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree v13.4 (d), resulting in the chronogram of summary source ages shown in (g).
The Paleogene Period spans from the end of the Cretaceous Period 66 million years ago (Ma) to the beginning of the Neogene Period 23.03 Ma.

Figure 3. Median summary chronogram resulting from a DateLife analysis of bird species within the family Fringillidae. For visualization purposes, we are showing a portion of the final median summary chronogram encompassing 57 species out of the 289 total included in the analysis. The complete final chronogram is available as Supplementary Figure S3. The starting tree topology (Supplementary Fig. S2) has 289 tip species and 253 nodes; DateLife revealed age data for 194 of these nodes from at least one published chronogram. In total, 19 different chronograms from 13 different studies contributed 818 age data points, which were summarized to obtain a single value for each one of the 194 nodes with age data. From the 194 summary ages available, 21 were discarded and not used as calibrations (asterisk, *), because they were older than a parent node or younger than a descendant node; the remaining 173 summary ages were used as secondary calibrations (forward slash, /) with the Branch Length Adjuster (BLADJ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}). The Paleogene Period spans from the end of the Cretaceous Period 66 million years ago (Ma) to the beginning of the Neogene Period 23.03 Ma.
